---
title: "semPlot practice"
author: "A Solomon Kurz"
date: "`r format(Sys.Date())`"
output:
  github_document
---

The purpose of this file is to practice working with the semPlot package.

Load the semPlot package.

```{r, warning = F, message = F}
# install.packages("semPlot", dependencies = T)
library(tidyverse)
library(semPlot)
library(lavaan)
```

Load the data.

```{r, warning = F, message = F}
socex1_names <- 
  c("w1vst1", "w1vst2", "w1vst3", "w2vst1", "w2vst2", "w2vst3", "w3vst1", "w3vst2", "w3vst3", "w1unw1", "w1unw2", "w1unw3", "w2unw1", "w2unw2", "w2unw3", "w3unw1", "w3unw2", "w3unw3", "w1dboth","w1dsad", "w1dblues", "w1ddep", "w2dboth", "w2dsad","w2dblues", "w2ddep", "w3dboth", "w3dsad", "w3dblues", "w3ddep", "w1marr2", "w1happy", "w1enjoy", "w1satis", "w1joyful", "w1please", "w2happy", "w2enjoy", "w2satis", "w2joyful", "w2please", "w3happy", "w3enjoy", "w3satis", "w3joyful", "w3please", "w1lea", "w2lea", "w3lea")

socex1 <- 
  read_table2("data/socex1.dat",
              col_names = F) %>% 
  set_names(socex1_names)
```

Here's our lavaan `model1.1`.

```{r}
model1.1 <- '
# loadings
w1comp =~ w1vst1 + w1vst2 + w1vst3
w2comp =~ w2vst1 + w2vst2 + w2vst3

# variances and covariances
w2comp ~~ w1comp
w1comp ~~ w1comp
w2comp ~~ w2comp
w1vst1 ~~ w1vst1
w1vst2 ~~ w1vst2
w1vst3 ~~ w1vst3
w2vst1 ~~ w2vst1
w2vst2 ~~ w2vst2
w2vst3 ~~ w2vst3
w1vst1 ~~ w2vst1
w1vst2 ~~ w2vst2
w1vst3 ~~ w2vst3
'

fitmodel1.1 <- sem(model1.1, data = socex1)
```

Here's the default semPlot figure via the `semPaths()` function.

```{r, fig.width = 6, fig.height = 3}
semPaths(fitmodel1.1)
```

Here's the same plot, but using `style = "lisrel"`.

```{r, fig.width = 6, fig.height = 3}
semPaths(fitmodel1.1,
         style = "lisrel")
```

You can rotate the diagram.

```{r, fig.width = 5, fig.height = 4}
semPaths(fitmodel1.1,
         style = "lisrel",
         rotation = 2)
```

You can also split up the layout of the measurement and structural models.

```{r, fig.width = 5, fig.height = 3}
semPaths(fitmodel1.1,
         style = "lisrel",
         layoutSplit = T,
         rotation = 2)
```

You might darken up the line color with the `edge.color` argument and globally change the fill color with `color`. While we at it, we can get rid of the shape outlines with `borders = F`.

```{r, fig.width = 5, fig.height = 3}
semPaths(fitmodel1.1,
         style = "lisrel",
         layoutSplit = T,
         rotation = 2,
         edge.color = "black",
         color = "grey85",
         borders = F)
```

In addition, you might change the size of the shapes and the font of their text, like so.

```{r, fig.width = 5, fig.height = 3}
semPaths(fitmodel1.1,
         style = "lisrel",
         layoutSplit = T,
         rotation = 2,
         edge.color = "black",
         color = "grey85",
         sizeMan = 7,
         sizeLat = 15,
         label.scale = T,
         label.cex = 1.5,
         borders = F)
```

If you want to relable any of the variables, you might index like using `$Arguments$labels`.
```{r, fig.width = 5, fig.height = 3}
semPaths(fitmodel1.1)$Arguments$labels
```

Once you have a sense of the order of the elements, you can then use `nodeLabels` to specify your updated labels. You can even use mathematical notation using the `expression` function.
```{r, fig.width = 5, fig.height = 3}
semPaths(fitmodel1.1,
         style = "lisrel",
         layoutSplit = T,
         rotation = 2,
         edge.color = "black",
         color = "grey85",
         sizeMan = 7,
         sizeLat = 15,
         label.scale = T,
         label.cex = 1.75,
         borders = F,
         nodeLabels = c(expression(y[1]), 
                        expression(y[2]),
                        expression(y[3]),
                        expression(y[4]),
                        expression(y[5]),
                        expression(y[6]),
                        expression(eta[1]),
                        expression(eta[2])))
```

With the `whatLabel` argument, we can add the parameter point estimates.

```{r, fig.width = 5, fig.height = 3}
semPaths(fitmodel1.1,
         style = "lisrel",
         layoutSplit = T,
         rotation = 2,
         edge.color = "black",
         color = "grey85",
         sizeMan = 7,
         sizeLat = 15,
         label.scale = T,
         label.cex = 1.75,
         borders = F,
         nodeLabels = c(expression(y[1]), 
                        expression(y[2]),
                        expression(y[3]),
                        expression(y[4]),
                        expression(y[5]),
                        expression(y[6]),
                        expression(eta[1]),
                        expression(eta[2])),
         whatLabel = "est")
```

```{r}
sessionInfo()
```

